---
title: "<br><br>Spatiotemporal modeling and predicting of average flows in Scotland"
subtitle: "based on generalized additive models"  
author: 
  - "Mengran Li"
output:
  xaringan::moon_reader:
    css: xaringan-themer.css
    nature:
      slideNumberFormat: "%current%"
      highlightStyle: github
      highlightLines: true
      ratio: 16:9
      countIncrementalSlides: true
    includes:
       in_header: "mathjax-equation-numbers.html"
header-includes:
- \usepackage{caption}
---

```{r setup, include=FALSE}
options(htmltools.dir.version = FALSE)
knitr::opts_chunk$set(
  fig.width=9, fig.height=3.5, fig.retina=3,
  out.width = "100%",
  cache = FALSE,
  echo = TRUE,
  message = FALSE, 
  warning = FALSE,
  hiline = TRUE
)
```

```{r xaringan-themer, include=FALSE, warning=FALSE}
library(xaringanthemer)
style_duo_accent(
  primary_color = "#00355F",
  secondary_color = "#FF961C",
  inverse_header_color = "#FFFFFF",
  background_image = "UoG_colour.png",
  background_size = "20%",
  background_position = "top 30px right 50px;",
  title_slide_background_image = "UoG_keyline.png",
  title_slide_background_size = "30%",
  title_slide_background_position = "50% 10%",
  title_slide_background_color = "#00355F",
  title_slide_text_color = "#FFFFFF",
  header_background_padding = "10px 10px 10px 0"
)
```

```{css, echo=FALSE}
pre {
  max-height: 300px;
  overflow-y: auto;
}

pre[class] {
  max-height: 100px;
}
```

```{css, echo=FALSE}
.scroll-4 {
  max-height: 4px;
  overflow-y: auto;
  background-color: inherit;
}
```

```{r, include=FALSE}
library(tidyverse)
library(maps)
library(geoR)
library(mgcv)
library(leaflet)
library(kableExtra)
library(cowplot)
library(broom)
library(mgcViz)
```

# Introduction

Research shows that the current climate is changing. One example is the trend of river flows. Understanding the temporal and spatial variation of river flow can help people reduce the loss of natural disasters such as flood and drought. This project will apply appropriate statistical methods to the river flows.

.pull-left[

## Aims of the project

- To detect the pattern of average daily flows **over space**.

- To detect the trend of average daily flows **over time**.

- To **predict** the average flows and **evaluate the efficiency** of the covariates in forecasting.
]

.pull-right[

## Data description

The daily flow data and gauge station information are provided by the **Scottish Environment Protection Agency (SEPA)** and the **National River Flow Archive(NRFA)**. 
]

---
# Data summarization

.pull-left[
The data contains observations of **65** gauge stations during **01/01/1989 - 31/12/2015**. The data of **29th Feb** are removed. Thus **N = 65 × 27 × 365 = 630720**, that is, **27** years of **365** daily observations in each station.

The obvious **positive skewed** distribution of average flow data is identified, therefore we take **logarithm** of flow. 

This dataset has **no missing values**.

```{r,echo = FALSE, fig.cap="Fig.1 Distribution of flow vs log(flow)."}
# Transform the Flow into log
load("files/data.Rdata")
data %>% select(Flow) %>% mutate(logFlow = log(Flow)) %>% gather(key = "group") %>%
  ggplot(aes(value)) +
  geom_density()+
  theme_xaringan()+
  facet_wrap(~group, scale = "free")
```
]

.pull-right[

```{r echo = FALSE, fig.width=4, fig.height=6}
Mean <- data %>% group_by(ID) %>% summarise(Mean = mean(logFlow))
UK <- map_data(map = "world", region = "UK") 

position <- data %>% select(ID, Longitude, Latitude)
position <- unique(position)
position <- position %>% left_join(Mean, c("ID" = "ID"))

pal <- colorNumeric(
  palette = "Blues",
  domain = position$Mean)

position %>% 
  leaflet() %>% 
  setView(lng = -4, lat = 57.2, zoom = 6.25) %>%
  addTiles(group = "OSM (default)") %>%
  addCircleMarkers(lng = ~Longitude,
                   lat = ~Latitude,
                   radius = ~ 4,
                   stroke = FALSE,
                   fill = TRUE,
                   fillOpacity =1,
                   color = "red",
                   labelOptions = labelOptions(noHide = T, textOnly =TRUE),
                   group = "Positions of gauging stations") %>%
  addCircleMarkers(data = position,
    lng = ~Longitude,
    lat = ~Latitude,
    radius = ~ 2*Mean,
    stroke = FALSE,
    fill = TRUE,
    color = ~ pal(Mean),
    fillOpacity =0.8,
    popup = ~as.character(ID),
    label = ~ as.character(ID),
    labelOptions = labelOptions(noHide = F, textOnly = TRUE),
    group = "Average Flow"
  ) %>%
  addLegend("bottomright", pal = pal, values = ~Mean,
    title = "Average Flow",
#    labFormat = labelFormat(prefix = "$"),
    opacity = 1, group = "Average Flow"
  ) %>%
  addLayersControl(
    baseGroups = c("Map"),
    overlayGroups = c("Positions of gauging stations", "Average Flow"),
    options = layersControlOptions(collapsed = FALSE)
  ) %>% 
   hideGroup("Average Flow")%>%
  addWMSTiles("https://basemap.nationalmap.gov/arcgis/services/USGSHydroCached/MapServer/WmsServer", layers = "0", group = "Map")
```

Distributions of gauging stations and average flow over space.

]

---
## Temporal


```{r, echo = FALSE,fig.cap = "Fig.2 Temporal exploratory data analysis, the trend plots of year, month, doy (date of year) and interaction between year and doy in A, B, C and D respectively."}
p1 <- data %>% 
  select("ID", "logFlow", "Year", "Month", "River") %>%
  group_by(Year) %>%
  summarise(MeanFlow = mean(logFlow)) %>%
  mutate(Year = as.numeric(Year)) %>%
  ggplot(aes(x=Year,y=MeanFlow, color = Year))+
  geom_line(color = "orange",size = 1)+
  geom_smooth(se = FALSE)+
#  ylim(c(-8,8)) +
  theme_xaringan(title_font_size =16,text_font_size =12)

p2 <- data %>% 
  select("ID", "logFlow", "Year", "Month", "River") %>%
  group_by(Month) %>%
  summarise(MeanFlow=mean(logFlow)) %>% 
  mutate(Month = as.numeric(Month)) %>%
  ggplot(aes(x=Month,y=MeanFlow))+
  geom_line(color = "orange",size = 1)+
  geom_smooth(se = FALSE)+
#  ylim(c(-8,8)) +
  theme_xaringan(title_font_size =16,text_font_size =12)

p3 <- data %>% 
  select("ID", "logFlow", "Year", "Month", "River", "doy") %>%
  group_by(doy) %>%
  summarise(MeanFlow=mean(logFlow)) %>% 
  ggplot(aes(x=doy,y=MeanFlow))+
  geom_smooth(se = FALSE)+
  geom_line(color = "orange",size = 1)+
#  ylim(c(-8,8)) +
  theme_xaringan(title_font_size =16,text_font_size =12)

p4 <- data %>% 
  select("ID", "logFlow", "Year", "doy", "River") %>%
  group_by(doy, Year) %>%
  summarise(MeanFlow=mean(logFlow)) %>% 
  ggplot(aes(x=doy,y=MeanFlow,group=Year, color = Year), color = Year)+
  geom_smooth(se = FALSE)+
  theme_xaringan(title_font_size =16,text_font_size =12)
plot_grid(p1,p2,p3,p4, ncol=2, labels="AUTO")
```

---
# Methodology

## Modelling

### Simple Linear model

$$ 
\begin{align}
   y_i = x_{1i}+x_{2i} +t_i + d_i + z_i
\end{align}
$$

### Modelling space and time seperately

$$ 
\begin{align}
   y_i = x_{1i}+x_{2i} +s_1(t_i) + s_2(d_i) + s_3(z_i)
\end{align}
$$

$$
\begin{align}
   y_i = x_{1i}+x_{2i}+s_1(t_i) + s_2(d_i) + s_3(z_i) + s_4(t_i,d_i)
\end{align}
$$

---

### Joint models of space and time
$$
\begin{align}
y_i = x_{1i}+x_{2i}+ s_1(t_i) + s_2(d_i) + s_3(z_i) + s_4(t_i,d_i) + s_5(t_i,z_i)+s_6(d_i,z_1)\label{eq:fit3}
\end{align}
$$

  where $y$ is **log(flow)**, $x_1$ and $x_2$ are **catchment area** and **max altitude**, $s_1(t)$ and $s_2(d)$ are smooth functions of **time (year)** and **doy (day of the year)**, and $s_3(z)$ is a bivariate smooth function of **coordinates (longitude, latitude)**.

## Model diagnosis and prediction

- Check the nested models to test the significance of covariates and smooth functions (**Chisq** test, **F** test)

- Select the best model (**AIC**, **BIC**, etc.)

- Residual correlations (**Q-Q plot**, **Spatial progress**)

- predictions (**Kriging**)

---
# Results

## Model Fitting

```{r, echo = FALSE}
load("files/fit.Rdata")
```

```{r, eval= FALSE, class.output="scroll-4"}
## linear model
fit <- lm(logFlow ~ Catchment.Area + Max.Altitude + 
            doy + Year + Longitude + Latitude, 
          data = data)
## Check the significance of smooth functions
# doy
smooth_doy <- gam(logFlow ~ Catchment.Area + Max.Altitude + 
              Year + Longitude + Latitude +
*             s(doy), 
            data = data)
# year
smooth_year <- gam(logFlow ~ Catchment.Area + Max.Altitude + 
              doy + Longitude + Latitude +
*              s(Year), 
            data = data)
# coordinates
smooth_coord <- gam(logFlow ~ Catchment.Area + Max.Altitude + 
              doy + Year + 
*             te(Longitude , Latitude), 
            data = data)
## Check interactions
# without interaction
gamfit1 <- gam(logFlow ~ Catchment.Area + Max.Altitude + 
                 s(doy) + s(Year) + te(Longitude, Latitude), 
               data = data)
# interaction of doy and year
gamfit2 <- gam(logFlow ~ Catchment.Area + Max.Altitude +
                 s(doy) + s(Year) + te(Longitude, Latitude) + 
*                ti(doy, Year), 
               data = data)
# all interactions
gamfit3 <- gam(logFlow ~ Catchment.Area + Max.Altitude +
                 s(doy) + s(Year) + te(Longitude, Latitude) +
                 ti(doy, Year)+
*                ti(Longitude, Latitude, doy, d=c(2,1)) +
*                ti(Longitude, Latitude, Year, d=c(2,1)), 
               data = data)
```

According to the Chisq test and F test, the covariates and smooth functions are all significant at 0.05. We compare the candidate three GAMs:

```{r, echo = FALSE, highlight.output=3}
f1 <- glance(gamfit1)
f2 <- glance(gamfit2)
f3 <- glance(gamfit3)
f <- rbind(f1,f2)
f <- rbind(f,f3)
f %>% add_column(Model = c("gamfit1","gamfit2","gamfit3"), .before = "df") %>% kable(format = 'html', caption = "Tab.1 Model comparison")
rm(f1,f2,f3,f)
```

The results display that the gamfit3 (Eq. \ref{eq:fit3}) is the best model.

---
## Estimated Parameters

```{r, echo = FALSE}
rm(p1, p2, p3, p4)
b <- getViz(gamfit3)
```


```{r, echo = FALSE, fig.cap="Fig.3 Estimated parameters of selected GAM."}
pl <- plot(b, allTerms = T) + 
  l_points() + 
  l_fitLine(linetype = 3) + 
  l_fitRaster()+
  l_fitContour() + 
  l_ciLine(colour = 2) + 
  l_ciBar() + 
  l_fitPoints(size = 1, col = 2) + 
  theme_get() + 
  labs(title = NULL)
print(pl, pages = 1)
```

---

## Residual correlations

```{r, echo=FALSE, results='hide'}
load("prediction.Rdata")
vari <- variog(residUK)
vari.mc <- variog.mc.env(residUK, obj.variog=vari)
```

.pull-left[

```{r, echo = FALSE, class.output="scroll-4"}
summary(gamfit3)
```

**Q-Q dialogue** plot indicates that there is **autocorrelation** among residuals. We check the spatial progress. 

]

.pull-right[
```{r, echo = FALSE, fig.cap = "Fig4. QQ plot of the residuals."}
qq(b, method = "simul1", a.qqpoi = list("shape" = 1), a.ablin = list("linetype" = 2))+
  theme_xaringan(title_font_size =16,text_font_size =12)
```

```{r, echo = FALSE, fig.cap = "Fig5. MC envelope plot."}
plot(vari, envelope.obj=vari.mc)
```

]

---
## Predictions

```{r, echo=FALSE, fig.cap = "Fig.6 Predicted average flow via kriging in Scotland in the day of 1 (1st Jan.), 92 (1st Apr.), 182 (1st Jul.) and 274 (1st Oct.) every 5 years.", fig.align='left'}
p2 + theme_bw()
```


---

# Conclusions

- For the **spatial** effect, the flow in the **surrounding** area is **higher** than that in the central area.  

- For the **temporal** effect, the average flow fluctuates by **5 years** and in **each year**, the mean flow **decreases** until **July** and **goes up** to the **end of the year**. With the **interaction** effect between time and season, the **highest** flow moves along with the **increasing of doy (day of the year) and year**.

- The next work is to **evaluate the efficiency** of the covariates in forecasting.

---
# References

```{r,results='asis', echo=FALSE}
library(RefManageR)
BibOptions(
check.entries = FALSE, 
bib.style = "authoryear", 
cite.style = "authoryear", 
style = "markdown",
hyperlink = FALSE, 
dashed = FALSE)
myBib = ReadBib("references.bib", check = FALSE)
SearchBib(myBib)
```
